---
title: "Template_Preliminary_Data_Exploration"
author: "Shanny Foo"
date: "26/06/2022"
output: word_document
---

#### HOUSECLEANING IN R ####
# To comment/uncomment out whole chunk: SHIFT+CTRL+C
# To Restart R: SHIFT+CTRL+F10
```{r Clearing R Workspace, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Clear plots
if(!is.null(dev.list())) dev.off()  # In GUI: clear all plots in the plots panel.

# Clear console
cat("\014")  # In GUI: Ctrl+L or to click on the clear console within the Edit menu.

# Clean workspace
rm(list=ls())  # In GUI: clear objects from the workspace in the environment panel.
```



--- SETUP

#### LIBRARIES ####
```{r warning=TRUE, include=FALSE}
#library(openxlsx)
library(readxl)

library(tidyverse)
library(ggpubr)
library(ggplot2)
```



--- INSPECT AND EXPLORE THE DATA

#### INSPECTION OF COL FREQUENCY DISTRIBUTION ####
```{r Plot col freq dist, echo=FALSE, message=FALSE, warning=TRUE, paged.print=FALSE}
### GET A SENSE OF COL DISTRIBUTION

## Look at the summary statistics of the variables
summary(Dataframe)


### VISUAL INSPECTION OF COL FREQUENCY DISTRIBUTION

## Single plot of data distribution facted by variable
NewDF <- reshape2::melt(Dataframe)

ggplot(NewDF, aes(value)) +
      facet_wrap(~variable, scales = 'free_x') +
      geom_histogram(binwidth = function(x) 2 * IQR(x) / (length(x)^(1/3)))
```


#### BOX PLOT: CATEGORICAL VARIABLE RANGES ####
```{r}
## Check the range and distribution of the DV by categorical variables

par(mar=c(1,1,1,1))  # set plot margins

# DV and IV1, IV2, IV3 ...
boxplot(DV ~ IV1, data = Dataframe)
boxplot(DV ~ IV2, data = Dataframe)
boxplot(DV ~ IV3, data = Dataframe)
```


#### SPAGHETTI PLOT: OVERALL DISTRIBUTION ####
```{r Plot overall dist, echo=FALSE, message=FALSE, warning=TRUE, paged.print=FALSE}
### Plot overall distribution of x and y variables (e.g., IV and DV) ###

## Label for plot title 
lab.DV = c("level1","level2","level3")  # Label for DV and sub-labels for each level of the DV


## Plot parameters
ggplot(data=Dataframe,
             aes(x=IV,y=DV,group=GroupingVariable)) +
               geom_line () + geom_point () + ggtitle("Plot Title") +
               theme_bw() +
               scale_x_continuous (breaks = c("level1","level2","level3"), labels=lab.DV, name = "DV") +
               scale_y_continuous (breaks = seq(-180,420,50), name= "IV") +
               stat_summary(fun=mean,geom="point",size=2.5,shape=18,colour="orange") +
               stat_summary(fun=mean,geom="line",lwd=0.4,colour="blue") +
               stat_smooth(method="lm",se=FALSE,lwd=0.2,formula=y~x,colour="red") +
               stat_smooth(method="lm",se=FALSE,lwd=0.2,formula=y~poly(x,2),colour="green")
```


#### SPAGHETTI PLOT: CONDITIONING ON MiniSEA-INVARIANT COVARIATES ####
```{r}
### Spaghetti plot with lowess smoother by grouping variable 1 to get a sense of the mean function by grouping variable 1 ###

## Set panels (row#, col#), starting position (1,1,1,1 from margin), adjusted r, and font subtitle
par(mfrow=c(3,1),oma=c(1,1,1,1),adj=0.5,font.sub=3) # note: for error that panels were out of bounds, decreased start position to c(0,0,0,0)


## Plot parameters for a group, with 3 conditions for subgroups
tmp1 = cbind(Dataframe$DV[Dataframe$GroupingVariable=="Condition1"],Dataframe$aINS[Dataframe$GroupingVariable=="Condition1"])

tmp2 = cbind(Dataframe$DV[Dataframe$GroupingVariable=="Condition2"],Dataframe$aINS[Dataframe$GroupingVariable=="Condition2"])

tmp3 = cbind(Dataframe$DV[Dataframe$GroupingVariable=="Condition3"],Dataframe$aINS[Dataframe$GroupingVariable=="Condition3"])

# EXample: GENFI data with 3 subgroups within clinical status (levles 0,1,2)
tmp1 = cbind(GENFI_Time_1$MiniSEA[GENFI_Time_1$ClinicalStatus=="0"],GENFI_Time_1$aINS[GENFI_Time_1$ClinicalStatus=="0"])

tmp2 = cbind(GENFI_Time_1$MiniSEA[GENFI_Time_1$ClinicalStatus=="1"],GENFI_Time_1$aINS[GENFI_Time_1$ClinicalStatus=="1"])

tmp3 = cbind(GENFI_Time_1$MiniSEA[GENFI_Time_1$ClinicalStatus=="2"],GENFI_Time_1$aINS[GENFI_Time_1$ClinicalStatus=="2"])


## Plot panels for each group, side by side, showing x~y relationship for each
scatter.smooth(tmp1[,1],tmp1[,2],pch=42,cex=0.75,xlab="DV",
               ylab="IV")
title("GroupingVariable = 1")

scatter.smooth(tmp1[,1],tmp1[,2],pch=42,cex=0.75,xlab="DV",
               ylab="IV")
title("GroupingVariable = 2")

scatter.smooth(tmp1[,1],tmp1[,2],pch=42,cex=0.75,xlab="DV",
               ylab="IV")
title("GroupingVariable = 3")

# Example: one panel for each subgroup of the clinical grouping variable showing relationship between MiniSEA and aINS
scatter.smooth(tmp1[,1],tmp1[,2],pch=42,cex=0.75,xlab="MiniSEA",
               ylab="aINS")
title("ClinicalGroup = NMC")

scatter.smooth(tmp2[,1],tmp2[,2],pch=42,cex=0.75,xlab="MiniSEA",
               ylab="aINS")
title("ClinicalGroup = PMC")

scatter.smooth(tmp3[,1],tmp3[,2],pch=42,cex=0.75,xlab="MiniSEA",
               ylab="aINS")
title("ClinicalGroup = SMC")
```


#### SCATTER PLOT: VARIABLE RELATIONSHIPS ####
```{r}
## Plot the relationship between DV and IV, facted by GroupingVariable1, without taking into consideration the multilevel structure of the data
ggplot(data  = Dataframe,
       aes(x = IV,
           y = IV))+
  geom_point(size = 1.2,
             alpha = .8,
             position = "jitter")+# to add some random noise for plotting purposes
  facet_wrap(~GroupingVariable1, nrow=4, ncol=1) + # number of rows should be same as number of conditiosn, one panel for each subgroup
  theme_minimal()+
  labs(title = "IV vs. DV")

# Regression line added to plot
ggplot(data  = Dataframe,
       aes(x = IV,
           y = DV))+
  geom_point(size     = 1.2,
             alpha    = .8,
             position = "jitter")+ #to add some random noise for plotting purposes
  geom_smooth(method = lm,
              se     = FALSE,
              col    = "black",
              size   = .5,
              alpha  = .8)+ # to add regression line
  facet_wrap(~ClinicalStatus, nrow=4, ncol=1) +  # number of rows should be same as number of conditiosn, one panel for each subgroup
  theme_minimal()+
  labs(title = "IV vs. DV")
  theme_minimal()+
  labs(title    = "X vs Y",
       subtitle = "with regression line")

# Separate regression lines for each GroupingVariable1 subgroup, faceted by GroupingVariable2
ggplot(data      = Dataframe,
       aes(x     = IV,
           y     = DV,
           col   = GroupingVariable1,
           group = GroupingVariable1))+ #to add the colours for different GroupingVariable1 subgroups
  geom_point(size     = 1.2,
             alpha    = .8,
             position = "jitter")+ #to add some random noise for plotting purposes
  facet_wrap(~GroupingVariable2, nrow=4, ncol=1) +  # number of rows should be same as number of conditiosn, one panel for each subgroup
  theme_minimal()+
  labs(title = "IV vs. DV")
  theme_minimal()+
  theme(legend.position = "top")+
  #theme(legend.position = "none")+
  scale_fill_steps2()+
  #scale_color_gradientn(colours = rainbow(100))+
  geom_smooth(method = lm,
              se     = FALSE,
              size   = .5,
              alpha  = .8)+ # to add regression line
  labs(title    = "X vs. Y",
       subtitle = "added colours for different GroupingVariable1 groups and associated regression lines")

##############################

# Example: GENFI data of MiniSEA~aINS relationship, faceted by ClinicalStatus
ggplot(data  = GENFI_Time_1,
       aes(x = MiniSEA,
           y = aINS))+
  geom_point(size = 1.2,
             alpha = .8,
             position = "jitter")+# to add some random noise for plotting purposes
  facet_wrap(~ClinicalStatus, nrow=4, ncol=1) +
  theme_minimal()+
  labs(title = "aINS vs. MiniSEA")

# Example: Regression of MiniSEA on aINS
ggplot(data  = GENFI_Time_1,
       aes(x = MiniSEA,
           y = aINS))+
  geom_point(size     = 1.2,
             alpha    = .8,
             position = "jitter")+ #to add some random noise for plotting purposes
  geom_smooth(method = lm,
              se     = FALSE,
              col    = "black",
              size   = .5,
              alpha  = .8)+ # to add regression line
  facet_wrap(~ClinicalStatus, nrow=4, ncol=1) +
  theme_minimal()+
  labs(title    = "aINS vs. MiniSEA",
       subtitle = "add regression line")

# Example: Separate regression lines for each ClinicalStatus subgroup, faceted by level of GeneticMutation group
ggplot(data      = GENFI_Time_1,
       aes(x     = MiniSEA,
           y     = aINS,
           col   = ClinicalStatus,
           group = ClinicalStatus))+ #to add the colours for different ClinicalStatus groups
  geom_point(size     = 1.2,
             alpha    = .8,
             position = "jitter")+ #to add some random noise for plotting purposes
  facet_wrap(~GeneticMutation, nrow=4, ncol=1) +
  theme_minimal()+
  theme(legend.position = "top")+
  #theme(legend.position = "none")+
  scale_fill_steps2()+
  #scale_color_gradientn(colours = rainbow(100))+
  geom_smooth(method = lm,
              se     = FALSE,
              size   = .5,
              alpha  = .8)+ # to add regression line
  labs(title    = "aINS vs. MiniSEA",
       subtitle = "add colours for different ClinicalStatus groups and regression lines")
```



























































